Plant Trait Imputation

### 2024-05-09 code/L1/L1_P4_TropicalAndes_plant_imputetraits.Rmd
get a single trait value for each family & genus, combine all family level trait data, combine all genus level trait data, add trait level column to dataframes, check column names for dataframes, combine non-imputed traits and family or genus level traits

# Seperate dataframes from list
Dispersal.Syndrome <- trait_dfs$Dispersal.Syndrome
Fruit.Length <- trait_dfs$Fruit.Length
Plant.Height <- trait_dfs$Plant.Height
Fruit.Type <- trait_dfs$Fruit.Type
Growth.Form <- trait_dfs$Growth.Form
```

@@

```{r}

@@

>>>>

# Separate dataframes from list
Dispersal.Syndrome <- trait_dfs$Dispersal.Syndrome
Fruit.Length <- trait_dfs$Fruit.Length
Plant.Height <- trait_dfs$Plant.Height
Fruit.Type <- trait_dfs$Fruit.Type
Growth.Form <- trait_dfs$Growth.Form
```

@@

Get a single trait value for each family/genus
```{r}
# numeric traits
numeric_traits_combined <- function(df, level) {
  df$trait_value <- as.numeric(df$trait_value)
  if (level == "genus") {
    df %>%
      group_by(scrubbed_genus) %>%
      summarise(trait_value = mean(trait_value, na.rm = TRUE)) %>%
      ungroup()
  } else if (level == "family") {
    df %>%
      group_by(scrubbed_family) %>%
      summarise(trait_value = mean(trait_value, na.rm = TRUE)) %>%
      ungroup()
  }
}

# non-numeric traits
cat_traits_combined <- function(df, level){
  if (level == "genus") {
    df %>%
      group_by(scrubbed_genus) %>%
      summarise_all(.funs = list(mode_trait = ~names(sort(table(.), decreasing = TRUE))[1])) %>%
      ungroup()
  } else if (level == "family") {
    df %>%
      group_by(scrubbed_family) %>%
      summarise_all(.funs = list(mode_trait = ~names(sort(table(.), decreasing = TRUE))[1])) %>%
      ungroup()
  }
}

@@

```{r}
Dispersal.Syndrome_families_BIEN_df <- cat_traits_combined(Dispersal.Syndrome_families_BIEN_df, "family")
Dispersal.Syndrome_genus_BIEN_df <- cat_traits_combined(Dispersal.Syndrome_genus_BIEN_df, "genus")
```

```{r}
Fruit.Length_families_BIEN_df <- numeric_traits_combined(Fruit.Length_families_BIEN_df, "family")
Fruit.Length_genus_BIEN_df <- numeric_traits_combined(Fruit.Length_genus_BIEN_df, "genus")
```

```{r}
Plant.Height_families_BIEN_df <- numeric_traits_combined(Plant.Height_families_BIEN_df, "family")
Plant.Height_genus_BIEN_df <- numeric_traits_combined(Plant.Height_genus_BIEN_df, "genus")
```

```{r}
Fruit.Type_families_BIEN_df <- cat_traits_combined(Fruit.Type_families_BIEN_df, "family")
Fruit.Type_genus_BIEN_df <- cat_traits_combined(Fruit.Type_genus_BIEN_df, "genus")
```

```{r}
Growth.Form_families_BIEN_df <- cat_traits_combined(Growth.Form_families_BIEN_df, "family")
Growth.Form_genus_BIEN_df <- cat_traits_combined(Growth.Form_genus_BIEN_df, "genus")
```

add column of trait level (specices, genus, family)
```{r}
# Add new column trait_level to long_plant_trait dataframe
long_plant_traits <- long_plant_traits %>%
  mutate(TraitLevel = ifelse(!is.na(TraitValue), "species", NA))
```

```{r}
# add family and genus information to long_plant_trait dataframe
long_plant_traits_tax <- merge(long_plant_traits, na_species_df, by = "species", all.x = TRUE)
```

```{r}
# combine all family level trait data
## add column of trait_name for numeric traits
Dispersal.Syndrome_families_BIEN_df$trait_name <- "Dispersal.Syndrome"
Fruit.Length_families_BIEN_df$trait_name <- "Fruit.Length_mm"
Plant.Height_families_BIEN_df$trait_name <- "Plant.Height_m"
Fruit.Type_families_BIEN_df$trait_name <- "Fruit.Type"
Growth.Form_families_BIEN_df$trait_name <- "Growth.Form"

## rename trait_value column so it's consistent with other traits
Fruit.Type_families_BIEN_df$trait_value <- Fruit.Type_families_BIEN_df$trait_value_mode_trait
Growth.Form_families_BIEN_df$trait_value <- Growth.Form_families_BIEN_df$trait_value_mode_trait

## convert numeric trait values into character
Plant.Height_families_BIEN_df$trait_value <- as.character(Plant.Height_families_BIEN_df$trait_value)
Fruit.Length_families_BIEN_df$trait_value <- as.character(Fruit.Length_families_BIEN_df$trait_value)

all_family_traits <- bind_rows(
  select(Dispersal.Syndrome_families_BIEN_df, scrubbed_family, trait_name, trait_value),
  select(Fruit.Length_families_BIEN_df, scrubbed_family, trait_name, trait_value),
  select(Plant.Height_families_BIEN_df, scrubbed_family, trait_name, trait_value),
  select(Fruit.Type_families_BIEN_df, scrubbed_family, trait_name, trait_value),
  select(Growth.Form_families_BIEN_df, scrubbed_family, trait_name, trait_value)
  )

## add trait level column
all_family_traits$TraitLevel <- "family"
```

```{r}
# combine all genus level trait data
## add column of trait_name for numeric traits
Dispersal.Syndrome_genus_BIEN_df$trait_name <- "Dispersal.Syndrome"
Fruit.Length_genus_BIEN_df$trait_name <- "Fruit.Length_mm"
Plant.Height_genus_BIEN_df$trait_name <- "Plant.Height_m"
Fruit.Type_genus_BIEN_df$trait_name <- "Fruit.Type"
Growth.Form_genus_BIEN_df$trait_name <- "Growth.Form"

## rename trait_value column so it's consistent with other traits
Fruit.Type_genus_BIEN_df$trait_value <- Fruit.Type_genus_BIEN_df$trait_value_mode_trait
Growth.Form_genus_BIEN_df$trait_value <- Growth.Form_genus_BIEN_df$trait_value_mode_trait

## convert numeric trait values into character
Plant.Height_genus_BIEN_df$trait_value <- as.character(Plant.Height_genus_BIEN_df$trait_value)
Fruit.Length_genus_BIEN_df$trait_value <- as.character(Fruit.Length_genus_BIEN_df$trait_value)

all_genus_traits <- bind_rows(
  select(Dispersal.Syndrome_genus_BIEN_df, scrubbed_genus, trait_name, trait_value),
  select(Fruit.Length_genus_BIEN_df, scrubbed_genus, trait_name, trait_value),
  select(Plant.Height_genus_BIEN_df, scrubbed_genus, trait_name, trait_value),
  select(Fruit.Type_genus_BIEN_df, scrubbed_genus, trait_name, trait_value),
  select(Growth.Form_genus_BIEN_df, scrubbed_genus, trait_name, trait_value)
  )

## add trait level column
all_genus_traits$TraitLevel <- "genus"
```

make sure column names match between dataframes before combining
```{r}
colnames(all_family_traits)
colnames(all_genus_traits)
colnames(long_plant_traits_tax)
```

```{r}
colnames(all_family_traits)[which(colnames(all_family_traits) == "scrubbed_family")] <- "family"
colnames(all_family_traits)[which(colnames(all_family_traits) == "trait_name")] <- "TraitName"
colnames(all_family_traits)[which(colnames(all_family_traits) == "trait_value")] <- "TraitValue"

colnames(all_genus_traits)[which(colnames(all_genus_traits) == "scrubbed_genus")] <- "genus"
colnames(all_genus_traits)[which(colnames(all_genus_traits) == "trait_name")] <- "TraitName"
colnames(all_genus_traits)[which(colnames(all_genus_traits) == "trait_value")] <- "TraitValue"
```

use family & genus level traits to fill in gaps of original dataframe
genus level prefered over family
```{r}
# Join the dataframes based on genus and TraitName
filled_long_plant_traits_tax <- long_plant_traits_tax %>%
  left_join(all_genus_traits, by = c("genus", "TraitName"), relationship = "many-to-many") %>%
  # Fill NAs in TraitValue column with values from all_genus_traits
  mutate(TraitValue = ifelse(is.na(TraitValue.x), TraitValue.y, TraitValue.x)) %>%
  # Remove the columns from all_genus_traits (if needed)
  select(-TraitValue.y)
```

Add units to wide trait table numeric trait names
fruit mass -> Fruit.Mass_mg
fruit length -> Fruit.Length_mm
plant height -> Plant.Height_m
plant lifespan -> Plant.Lifespan_years
seed mass -> Seed.Mass_mg
seed length -> Seed.Length_mm
seed width -> Seed.Width_mm
